Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-07-13 appendix-EPQ.tex

Let us begin to represent register automata that is used for capture the existence of a data-differentiated execution $e$, let $p$ be a maximal priority of $e$, and $e'$ be the projection of $e$ into values with priorities comparable with $p$. We also require that $\mathsf{Has\text{-}MatchedMaxPriority}(e')$ holds and there exists a cycle going through the value with maximal priority in $e'$. By data-independence, we can obtain $e_r$ from $e$ by renaming function, which maps such value to be $b$, maps values that cover it to be $a$, and maps other values into $d$. There are four possible enumeration of call and return actions of $\textit{put}(b,p)$ and $\textit{rm}(b)$. For each of them, we generate a register automaton.

For the case when $e_r \vert_{b} = \textit{call}(\textit{put},b,p) \cdot \textit{ret}(\textit{put},b,p) \cdot \textit{call}(\textit{rm},b) \cdot \textit{ret}(\textit{rm},b)$, we generate register automaton $\mathcal{A}_{\textit{l-lar}}^1$, as shown in \figurename~\ref{fig:automata APQ1Lar-1}. Here $C_1 = C \cup \{ \textit{ret}(\textit{rm},a) \}$, $C_2 = C \cup \{ \textit{call}(\textit{put},a,<r) \}$, $C_3 = C_2 \cup \{ \textit{ret}(\textit{rm},a) \}$, where $C = \{ \textit{call}(\textit{put},\top,\textit{true}),\textit{ret}(\textit{put},\top,\textit{true}), \textit{call}(\textit{rm},\top)$, $\textit{ret}(\textit{rm},\top),$ $\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty})$. The differentiated branch in $\mathcal{A}_{\textit{l-lar}}^1$ comes from the positions of the first $\textit{ret}(\textit{put},a,\_)$.

@@

For the case when $e_r \vert_{b} = \textit{call}(\textit{put},b,p) \cdot \textit{call}(\textit{rm},b) \cdot \textit{ret}(\textit{put},b,p) \cdot \textit{ret}(\textit{rm},b)$, we generate register automaton $\mathcal{A}_{\textit{l-lar}}^2$, as shown in \figurename~\ref{fig:automata APQ1Lar-2}. Here $C_1,C_2,C_3$ is the same as that in $\mathcal{A}_{\textit{l-lar}}^1$. The differentiated branch in $\mathcal{A}_{\textit{l-lar}}^2$ comes from the positions of the first $\textit{ret}(\textit{put},a,\_)$.

@@

For the case when $e_r \vert_{b} = \textit{call}(\textit{rm},b) \cdot \textit{call}(\textit{put},b,p) \cdot \textit{ret}(\textit{put},b,p) \cdot \textit{ret}(\textit{rm},b)$, we generate register automaton $\mathcal{A}_{\textit{l-lar}}^3$, as shown in \figurename~\ref{fig:automata APQ1Lar-3}. Here $C_1,C_2,C_3$ is the same as that in $\mathcal{A}_{\textit{l-lar}}^1$. The differentiated branch in $\mathcal{A}_{\textit{l-lar}}^3$ comes from the positions of the first $\textit{ret}(\textit{put},a,\_)$.

@@

For the case when $e_r \vert_{b} = \textit{call}(\textit{rm},b) \cdot \textit{call}(\textit{put},b,p) \cdot \textit{ret}(\textit{rm},b) \cdot \textit{ret}(\textit{put},b,p)$, we generate register automaton $\mathcal{A}_{\textit{l-lar}}^4$, as shown in \figurename~\ref{fig:automata APQ1Lar-4}. Here $C_1,C_2,C_3$ is the same as that in $\mathcal{A}_{\textit{l-lar}}^1$, and $C_4 = C_1 \cup \{ \textit{ret}(\textit{put},b,=r) \}$. The differentiated branch in $\mathcal{A}_{\textit{l-lar}}^4$ comes from the positions of the first $\textit{ret}(\textit{put},a,\_)$.

@@

\noindent The $\textit{if}$ direction: Given such $e$, $e'$ and $x$. Let renaming function $r$ maps $x$ into $b$, maps values cover $x$ into $a$, and maps other values into $d$. By data-independence, $r(e) \in \mathcal{I}$. Then depending on the cases of $r(e) \vert_{b}$, we can see that $r(e)$ is accepted by $\mathcal{A}_{\textit{l-lar}}^1$, $\mathcal{A}_{\textit{l-lar}}^2$, $\mathcal{A}_{\textit{l-lar}}^3$ or $\mathcal{A}_{\textit{l-lar}}^4$. \qed
\end {proof}

@@

An automaton $\mathcal{A}_{\textit{l-eq}}^1$ is given in \figurename~\ref{fig:automata for first enumeration of PQ1Equal}, and it is constructed for the first enumeration in \figurename~\ref{fig:history enumeration 1 for PQ1Equal}. Here we rename the values that cover $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},b)$ into $d$, and rename the remanning values into $\top$. In this figure, $C = \{ \textit{call}(\textit{put},\top,\textit{true}),\textit{ret}(\textit{put},\top,\textit{true})$, $\textit{call}(\textit{rm},\top), \textit{ret}(\textit{rm},\top),$ $\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty}) \}$, $C_1 = C \cup \{ \textit{call}(\textit{put},d,<r) \}$, $C_2 = C_1 \cup \{ \textit{ret}(\textit{put},b,=r) \}$, $C_3 = C_2 \cup \{ \textit{call}(\textit{put},d,<r),\textit{ret}(\textit{rm},d) \}$, $C_4 = C \cup \{ \textit{ret}(\textit{put},b,=r), \textit{ret}(\textit{rm},d) \}$.

@@

An automaton $\mathcal{A}_{\textit{l-eq}}^2$ is given in \figurename~\ref{fig:automata for second enumeration of PQ1Equal}, and it is constructed for the second enumeration in \figurename~\ref{fig:history enumeration 2 for PQ1Equal}. In \figurename~\ref{fig:automata for second enumeration of PQ1Equal}, $C_1$, $C_2$, $C_3$ and $C_4$ is same as that in \figurename~\ref{fig:automata for first enumeration of PQ1Equal}.

@@

Therefore, there are fourteen possible cases that satisfy the third enumeration in \figurename~\ref{fig:history enumeration 3 for PQ1Equal}. For each case, we construct an finite automaton. Let $\mathcal{A}_{\textit{1-eq}}^{3}$ be the union of register automata that is constructed for above fourteen cases. For example, for the case $\textit{ca}_1$ when $\textit{call}(\textit{put},a,p)$ is before $\textit{call}(\textit{rm},a_1)$, $\textit{call}(\textit{put},b,p)$ is after $\textit{ret}(\textit{rm},a_1)$, and $\textit{call}(\textit{put},b,p)$ is before $\textit{call}(\textit{rm},b)$, we construct a finite automaton $\mathcal{A}_{\textit{l-eq}}^{\textit{3-1}}$ in \figurename~\ref{fig:automata for ca1 of third enumeration of Rpr2}. In \figurename~\ref{fig:automata for ca1 of third enumeration of Rpr2}, let $C$ and $C_1$ the same as that in \figurename~\ref{fig:automata for first enumeration of PQ1Equal}. Let $C_2 = C_1 \cup \{ \textit{ret}(\textit{put},a_1,=r) \}$, $C_3 = C_2 \cup \{ \textit{ret}(\textit{put},b,=r) \}$, $C_4 = C_3 \cup \{ \textit{call}(\textit{put},d,<r), \textit{ret}(\textit{rm},d) \}$, and $C_5 = C \cup \{ \textit{ret}(\textit{put},b,=r), \textit{ret}(\textit{put},a_1,=r), \textit{ret}(\textit{rm},d) \}$. Other register automata can be similarly constructed.

@@

\noindent The $\textit{only if}$ direction: Assume that $e_1 \in \mathcal{I}$ is accepted by some register automata in $\mathcal{A}_{\textit{1-eq}}$. By data-independence, there exists data-differentiated execution $e_2 \in \mathcal{I}$ and a renaming function $r$, such that $e_1=r(e_2)$. Since $e_1$ is accepted by some register automata in  $\mathcal{A}_{\textit{1-eq}}$, let $x$, $y$ and $z$ (if exists) be the values that are renamed into $b$, $a$ and $a_1$ (if exists) by $r$, respectively, and let $d_1,\ldots,d_m$ be the values that are renamed into $d$ by $r$.

@@

\noindent The $\textit{if}$ direction: Assume that there exists $e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, such that $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e')$ holds $e'$ has only one maximal priority, $a'$ and $b'$ are two values with maximal priority $p$ in $e'$, $a' <_{\textit{pb}}^* b'$ in $e'$, and the rightmost gap-point of $b'$ is before $\textit{call}(\textit{put},a',p)$ or $\textit{call}(\textit{rm},a')$ in $e'$. By data-independence, we can obtain execution $e_1$ as follows: (1) rename $a'$ and $b'$ into $a$ and $b$, respectively, (2) for the values $d_1,\ldots,d_m$ that constitute the rightmost gap of $b'$, we rename them into $d$, (3) if $a' <_{\textit{pb}}^A a'_1 <_{\textit{pb}}^B b$, we rename $a'_1$ into $a_1$, and (4) rename the other values into $\top$. It is easy to see that $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e_1)$ holds, $a$ and $b$ has maximal priority in $e_1$, $a <_{\textit{pb}}^* b$ in $e_1$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,p)$ or $\textit{call}(\textit{rm},a)$ in $e_1$. By Lemma \ref{lemma:five enumeration is enough for EPQ1Equal}, there are five possible enumeration of operations of $a$, $b$, $a_1$ (if exists). Then

@@

Let us begin to represent a register automaton that is used for capture the case that, in a sub-execution $e'$ of an execution $e$, $\mathsf{Has\text{-}EmptyRemove}(e')$ holds, $e'$ is not $\mathsf{EmptyRemove}$-linearizable, and the reason is that there is a cycle going through some $\textit{rm}(\textit{empty})$ $o$ in the left-right constraint of $o$. The automaton is $\mathcal{A}_{\seqPQ}^3$, which is given in \figurename~\ref{fig:automata for PQ3}. In \figurename~\ref{fig:automata for PQ3}, let $C = \{ \textit{call}(\textit{put},\top,\textit{true}),\textit{ret}($ $\textit{put},\top,\textit{true}), \textit{call}(\textit{rm},\top), \textit{ret}(\textit{rm},\top),\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty}) \}$, $C_1 = C \cup \{ \textit{call}(\textit{put},b,\textit{true}) \}$, $C_2 = C_1 \cup \{ \textit{ret}(\textit{rm},b) \}$, and $C_3 = C \cup \{ \textit{ret}(\textit{rm},b) \}$.

@@

\noindent {\bf $\textit{fact}_1$}: Given a data-independent implementation $\mathcal{I}$. $\mathcal{A}_{\seqPQ}^3 \cap \mathcal{I} \neq \emptyset$ if and only if $\exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, $\mathsf{Has\text{-}EmptyRemove}(e')$ holds, $o = \textit{rm}(\textit{empty})$ is in $e'$, and $o$ is covered by some values $d_1,\ldots,d_m$ in $e'$.

\noindent The $\textit{only if}$ direction: Assume that $e_1 \in \mathcal{I}$ is accepted by $\mathcal{A}_{\seqPQ}^3$. By data-independence, there exists data-differentiated execution $e_2 \in \mathcal{I}$ and a renaming function $r$, such that $e_1=r(e_2)$. Let $d_1,\ldots,d_m$ be the values in $e_2$ such that $r(d_i)=b$ for each $1 \leq i \leq m$. Let $e_3 = e_2 \vert_{ \{ o, d_1, \ldots, d_m \} }$. It is obvious that $e_3 \in \textit{proj}(e_2)$ and $\textit{last}(e_3) = \seqPQ_3$. It is easy to see that $o$ is covered by $d_1,\ldots,d_m$.

\noindent The $\textit{if}$ direction: Assume that there exists such $e$, $e'$, $o$ and $d_1,\ldots,d_m$. Then, let $e_1$ be obtained from $e$ by renaming $d_1,\ldots,d_m$ into $b$ and renaming other values into $\top$. By data-independence, $e_1 \in \mathcal{I}$. It is easy to see that $e_1$ is accepted by $\mathcal{A}_{\seqPQ}^3$.

@@

When considering decidability and complexity, we consider only finite variables with finite data domain. In such case, the priority queue implementations can only model priority queue with bounded capacity. Then, at some time, the priority queue may become full and then put can not succeed until some item is removed. However, we assume that each method can always make progress in isolation, which means that $\textit{put}$ must return some value when the priority queue is full. Therefore, we introduce a specific value $\textit{full}$. A $\textit{put}$ method returns $\textit{full}$ indicates that the priority queue is full now. To fit our results of priority queue in former section, when checking linearizability of priority queue, we only consider executions without any $\textit{put}(\_,\_,\textit{full})$. Since does not mean we consider only executions with bounded number of operations, since an execution can contain unbounded number of pairs of put and remove operations by linearizable w.r.t pairs of put and remove operations.

@@

\noindent {\bf Reducing priority Queue into PSPACE and EXPSPACE problems:} We describe a class $\mathcal{C}$ of data-independent implementations for which linearizability w.r.t. $\seqPQ$ is decidable. The implementations in $\mathcal{C}$ allow an unbounded number of values but a bounded number of priorities. Each method manipulates a finite number of local variables which store Boolean values, or data values from $\mathbb{D}$. Methods communicate through a finite number of shared variables that also store Boolean values, or data values from $\mathbb{D}$. Data values may be assigned, but never used in program predicates (e.g., in the conditions of if-then-else and while statements) so as to ensure data independence. This class captures typical implementations, or finite-state abstractions thereof, e.g., obtained via predicate abstraction. Since the $\Gamma$-complete automata $A(\Gamma)$ uses a fixed set $D=\{a,b,a_1,d,e,\top\}$ of values, we have that $\mathcal{C}\cap A(\Gamma)\neq\emptyset$ for some $\Gamma$ iff $\mathcal{C}_D\cap A(\Gamma)\neq\emptyset$ where $\mathcal{C}_D$ is the subset of $\mathcal{C}$ that uses only values in $D$.

We model the set $\mathcal{C}_D$ as the executions of a Vector Addition System with States (VASS) similarly to Bouajjani et al. \cite{conf/esop/BouajjaniEEH13}. A VASS has a finite set of states, and manipulates a finite set of counters holding non-negative values. The idea is that the states of the VASS represent the global variables of $\mathcal{C}_D$. This set is finite as we have bounded the data values. Each counter of the VASS then represents the number of threads which are at a particular control location within a method, with a certain valuation of the local variables. Here again, as the data values are bounded, there is a finite number of valuations. When a thread moves from a control location to another, or updates its local variables, we decrement, resp., increment, the counter corresponding the old, resp., the new, control location or valuation.

For a fixed set of priorities $P$, the register automata $A(\Gamma)$ can be transformed to finite-state automata since the number of possible valuations of the registers is bounded. In this way, we reduce checking linearizability of priority queue into the the EXPSPACE-complete problem of checking state-reachability in a VASS. When we consider only finite number of threads, the value of counters are bounded and then this problem is PSPACE \cite{DBLP:conf/ac/Esparza96}.

\noindent {\bf Reducing PSPACE and EXPSPACE problems into linearizability of priority queue for bounded number of threads:} We show how to reduce the state-reachability problem of VASS into checking linearizability of priority queue. Our construction is based on the reduction of state-reachability problem of counter machine and VASS into linearizability in \cite{conf/esop/BouajjaniEEH13}.

There are already several linearizable lock-free implementations of priority queue \cite{DBLP:conf/ipps/SundellT03,DBLP:conf/ppopp/LiuS12}. It is not hard to modify them to fit finite variables with finite data domain and it is obvious that they are still linearizable then. Let $\textit{Imp-PQ}_{\textit{lf}}$ be one such implementations. It is safe to assume that $\textit{Imp-PQ}_{\textit{lf}}$ is given in the form of pseudo-code. We use $\textit{put}_{\textit{lf}}$ and $\textit{rm}_{\textit{lf}}$ to explicitly denote the pseudo-code of put and remove method of $\textit{Imp-PQ}_{\textit{lf}}$.

@@

>>>>

Let us begin to represent register automata that is used for capture the existence of a data-differentiated execution $e$, let $p$ be a maximal priority of $e$, and $e'$ be the projection of $e$ into values with priorities comparable with $p$. We also require that $\mathsf{Has\text{-}MatchedMaxPriority}(e')$ holds and there exists a cycle going through the value with maximal priority in $e'$. By data-independence, we can obtain $e_r$ from $e$ by renaming function, which maps such value to be $b$, maps values that cover it to be $a$, and maps other values into $\top$. There are four possible enumeration of call and return actions of $\textit{put}(b,p)$ and $\textit{rm}(b)$. For each of them, we generate a register automaton.

For the case when $e_r \vert_{b} = \textit{call}(\textit{put},b,p) \cdot \textit{ret}(\textit{put},b,\textit{true}) \cdot \textit{call}(\textit{rm},b) \cdot \textit{ret}(\textit{rm},b)$, we generate register automaton $\mathcal{A}_{\textit{l-lar}}^1$, as shown in \figurename~\ref{fig:automata APQ1Lar-1}. Here $A_1 = A \cup \{ \textit{ret}(\textit{rm},a) \}$, $A_2 = A \cup \{ \textit{call}(\textit{put},a,<r) \}$, $A_3 = A_2 \cup \{ \textit{ret}(\textit{rm},a) \}$, where $A = \{ \textit{call}(\textit{put},\top,\textit{true}),\textit{ret}(\textit{put},\top,\textit{true}), \textit{call}(\textit{rm},\top)$, $\textit{ret}(\textit{rm},\top),$ $\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty})$. The differentiated branch in $\mathcal{A}_{\textit{l-lar}}^1$ comes from the positions of the first $\textit{ret}(\textit{put},a,\_)$.

@@

For the case when $e_r \vert_{b} = \textit{call}(\textit{put},b,p) \cdot \textit{call}(\textit{rm},b) \cdot \textit{ret}(\textit{put},b,\textit{true}) \cdot \textit{ret}(\textit{rm},b)$, we generate register automaton $\mathcal{A}_{\textit{l-lar}}^2$, as shown in \figurename~\ref{fig:automata APQ1Lar-2}. Here $A_1,A_2,A_3$ is the same as that in $\mathcal{A}_{\textit{l-lar}}^1$. The differentiated branch in $\mathcal{A}_{\textit{l-lar}}^2$ comes from the positions of the first $\textit{ret}(\textit{put},a,\_)$.

@@

For the case when $e_r \vert_{b} = \textit{call}(\textit{rm},b) \cdot \textit{call}(\textit{put},b,p) \cdot \textit{ret}(\textit{put},b,\textit{true}) \cdot \textit{ret}(\textit{rm},b)$, we generate register automaton $\mathcal{A}_{\textit{l-lar}}^3$, as shown in \figurename~\ref{fig:automata APQ1Lar-3}. Here $A_1,A_2,A_3$ is the same as that in $\mathcal{A}_{\textit{l-lar}}^1$. The differentiated branch in $\mathcal{A}_{\textit{l-lar}}^3$ comes from the positions of the first $\textit{ret}(\textit{put},a,\_)$.

@@

For the case when $e_r \vert_{b} = \textit{call}(\textit{rm},b) \cdot \textit{call}(\textit{put},b,p) \cdot \textit{ret}(\textit{rm},b) \cdot \textit{ret}(\textit{put},b,\textit{true})$, we generate register automaton $\mathcal{A}_{\textit{l-lar}}^4$, as shown in \figurename~\ref{fig:automata APQ1Lar-4}. Here $A_1,A_2,A_3$ is the same as that in $\mathcal{A}_{\textit{l-lar}}^1$, and $A_4 = A_1 \cup \{ \textit{ret}(\textit{put},b,\textit{true}) \}$. The differentiated branch in $\mathcal{A}_{\textit{l-lar}}^4$ comes from the positions of the first $\textit{ret}(\textit{put},a,\_)$.

@@

\noindent The $\textit{if}$ direction: Given such $e$, $e'$ and $x$. Let renaming function $r$ maps $x$ into $b$, maps values cover $x$ into $a$, and maps other values into $\top$. By data-independence, $r(e) \in \mathcal{I}$. Then depending on the cases of $r(e) \vert_{b}$, we can see that $r(e)$ is accepted by $\mathcal{A}_{\textit{l-lar}}^1$, $\mathcal{A}_{\textit{l-lar}}^2$, $\mathcal{A}_{\textit{l-lar}}^3$ or $\mathcal{A}_{\textit{l-lar}}^4$. \qed
\end {proof}

@@

An automaton $\mathcal{A}_{\textit{l-eq}}^1$ is given in \figurename~\ref{fig:automata for first enumeration of PQ1Equal}, and it is constructed for the first enumeration in \figurename~\ref{fig:history enumeration 1 for PQ1Equal}. Here we rename the values that cover $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},b)$ into $c$, and rename the remanning values into $\top$. In this figure, $A = \{ \textit{call}(\textit{put},\top,\textit{true}),\textit{ret}(\textit{put},\top,\textit{true})$, $\textit{call}(\textit{rm},\top), \textit{ret}(\textit{rm},\top),$ $\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty}) \}$, $A_1 = A \cup \{ \textit{call}(\textit{put},c,<r) \}$, $A_2 = A_1 \cup \{ \textit{ret}(\textit{put},b,\textit{true}) \}$, $A_3 = A_2 \cup \{ \textit{ret}(\textit{rm},c) \}$, $A_4 = A \cup \{ \textit{ret}(\textit{put},b,\textit{true}), \textit{ret}(\textit{rm},c) \}$.

@@

An automaton $\mathcal{A}_{\textit{l-eq}}^2$ is given in \figurename~\ref{fig:automata for second enumeration of PQ1Equal}, and it is constructed for the second enumeration in \figurename~\ref{fig:history enumeration 2 for PQ1Equal}. In \figurename~\ref{fig:automata for second enumeration of PQ1Equal}, $A_1$, $A_2$, $A_3$ and $A_4$ is same as that in \figurename~\ref{fig:automata for first enumeration of PQ1Equal}.

@@

Therefore, there are fourteen possible cases that satisfy the third enumeration in \figurename~\ref{fig:history enumeration 3 for PQ1Equal}. For each case, we construct an finite automaton. Let $\mathcal{A}_{\textit{1-eq}}^{3}$ be the union of register automata that is constructed for above fourteen cases. For example, for the case $\textit{ca}_1$ when $\textit{call}(\textit{put},a,p)$ is before $\textit{call}(\textit{rm},a_1)$, $\textit{call}(\textit{put},b,p)$ is after $\textit{ret}(\textit{rm},a_1)$, and $\textit{call}(\textit{put},b,p)$ is before $\textit{call}(\textit{rm},b)$, we construct a finite automaton $\mathcal{A}_{\textit{l-eq}}^{\textit{3-1}}$ in \figurename~\ref{fig:automata for ca1 of third enumeration of Rpr2}. In \figurename~\ref{fig:automata for ca1 of third enumeration of Rpr2}, let $A$ and $A_1$ the same as that in \figurename~\ref{fig:automata for first enumeration of PQ1Equal}. Let $A_2 = A_1 \cup \{ \textit{ret}(\textit{put},a_1,\textit{true}) \}$, $A_3 = A_2 \cup \{ \textit{ret}(\textit{put},b,\textit{true}) \}$, $A_4 = A_3 \cup \{ \textit{ret}(\textit{rm},c) \}$, and $A_5 = A \cup \{ \textit{ret}(\textit{put},b,\textit{true}), \textit{ret}(\textit{put},a_1,\textit{true}), \textit{ret}(\textit{rm},c) \}$. Other register automata can be similarly constructed.

@@

\noindent The $\textit{only if}$ direction: Assume that $e_1 \in \mathcal{I}$ is accepted by some register automata in $\mathcal{A}_{\textit{1-eq}}$. By data-independence, there exists data-differentiated execution $e_2 \in \mathcal{I}$ and a renaming function $r$, such that $e_1=r(e_2)$. Since $e_1$ is accepted by some register automata in  $\mathcal{A}_{\textit{1-eq}}$, let $x$, $y$ and $z$ (if exists) be the values that are renamed into $b$, $a$ and $a_1$ (if exists) by $r$, respectively, and let $d_1,\ldots,d_m$ be the values that are renamed into $c$ by $r$.

@@

\noindent The $\textit{if}$ direction: Assume that there exists $e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, such that $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e')$ holds $e'$ has only one maximal priority, $a'$ and $b'$ are two values with maximal priority $p$ in $e'$, $a' <_{\textit{pb}}^* b'$ in $e'$, and the rightmost gap-point of $b'$ is before $\textit{call}(\textit{put},a',p)$ or $\textit{call}(\textit{rm},a')$ in $e'$. By data-independence, we can obtain execution $e_1$ as follows: (1) rename $a'$ and $b'$ into $a$ and $b$, respectively, (2) for the values $d_1,\ldots,d_m$ that constitute the rightmost gap of $b'$, we rename them into $c$, (3) if $a' <_{\textit{pb}}^A a'_1 <_{\textit{pb}}^B b'$, we rename $a'_1$ into $a_1$, and (4) rename the other values into $\top$. It is easy to see that $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e_1)$ holds, $a$ and $b$ has maximal priority in $e_1$, $a <_{\textit{pb}}^* b$ in $e_1$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,p)$ or $\textit{call}(\textit{rm},a)$ in $e_1$. By Lemma \ref{lemma:five enumeration is enough for EPQ1Equal}, there are five possible enumeration of operations of $a$, $b$, $a_1$ (if exists). Then

@@

Let us begin to represent a register automaton that is used for capture the case that, in a sub-execution $e'$ of an execution $e$, $\mathsf{Has\text{-}EmptyRemove}(e')$ holds, $e'$ is not $\mathsf{EmptyRemove}$-linearizable, and the reason is that there is a cycle going through some $\textit{rm}(\textit{empty})$ $o$ in the left-right constraint of $o$. The automaton is $\mathcal{A}_{\seqPQ}^3$, which is given in \figurename~\ref{fig:automata for PQ3}. In \figurename~\ref{fig:automata for PQ3}, let $A = \{ \textit{call}(\textit{put},\top,\textit{true}),\textit{ret}($ $\textit{put},\top,\textit{true}), \textit{call}(\textit{rm},\top), \textit{ret}(\textit{rm},\top),\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty}) \}$, $A_1 = A \cup \{ \textit{call}(\textit{put},b,\textit{true}) \}$, $C_2 = A_1 \cup \{ \textit{ret}(\textit{rm},b) \}$, and $A_3 = A \cup \{ \textit{ret}(\textit{rm},b) \}$.

@@

\noindent {\bf $\textit{fact}_1$}: Given a data-independent implementation $\mathcal{I}$. $\mathcal{A}_{\seqPQ}^3 \cap \mathcal{I} \neq \emptyset$, if and only if there exists $o = \textit{rm}(\textit{empty})$ in $e \in \mathcal{I}$, $G$ has a cycle going through $o$, where $G$ is the graph representing the left-right constraint of $o$.

\noindent The $\textit{only if}$ direction: Assume that $e_1 \in \mathcal{I}$ is accepted by $\mathcal{A}_{\seqPQ}^3$. By data-independence, there exists data-differentiated execution $e_2 \in \mathcal{I}$ and a renaming function $r$, such that $e_1=r(e_2)$. Since $e_1$ is accepted by $\mathcal{A}_{\seqPQ}^3$, there exists an operation $o=\textit{rm}(\textit{empty})$ in $e_1$ and $e_2$, and let $d_1,\ldots,d_m$ be the values that are renamed into $b$ by $r$. Let $e'' = e_2 \vert_{ \{ o,d_1,\ldots,d_m \} }$. It is not hard to see that $G$ has a cycle going through $o$, where $G$ is the graph representing the left-right constraint of $o$.

\noindent The $\textit{if}$ direction: Given such $e$ and $o$. By data-independence, there exists data-differentiated execution $e_1 \in \mathcal{I}$ and a renaming function $r_1$, such that $e=r_1(e_1)$. It is not hard to see that in $e_1$, $o$ is covered by values $d_1,\ldots,d_m$. Let renaming function $r_2$ be a renaming function which maps $d_1,\ldots,d_m$ into $b$ and all other values into $\top$. Then it is easy to see that $r_2(e_1) \in \mathcal{I}$ is accepted by $\mathcal{A}_{\seqPQ}^3$.

@@

When considering decidability and complexity, we consider only finite variables with finite data domain. In such case, it seems that the priority queue implementations can only model priority queue with bounded capacity. Then, at some time, the priority queue may become full and then put can not succeed until some item is removed. However, we assume that each method can always make progress in isolation, which means that $\textit{put}$ must return some value when the priority queue is full. Therefore, we introduce a specific value $\textit{full}$. A $\textit{put}$ method returns $\textit{full}$ indicates that the priority queue is full now. To fit our results of priority queue in former section, when checking linearizability of priority queue, we only consider executions without any $\textit{put}(\_,\_,\textit{full})$. This does not mean we consider only executions with bounded number of operations, since an execution can contain unbounded number of pairs of put and remove operations by linearizable w.r.t pairs of put and remove operations.

@@

\noindent {\bf Reducing Linearizability Problem of Priority Queue into State-Reachability Problem of VASS:} We describe a class $\mathcal{C}$ of data-independent implementations for which linearizability w.r.t. $\seqPQ$ is decidable. The implementations in $\mathcal{C}$ allow an unbounded number of values but a bounded number of priorities. Each method has a finite set of local variables storing Boolean values or values from $\mathbb{D}$. Methods communicate through a finite number of shared variables interpreted also as Booleans or values from $\mathbb{D}$. To ensure data independence, values in $\mathbb{D}$ may be assigned, but never used in Boolean expressions (e.g., of if-then-else statements). This class captures typical implementations, or finite-state abstractions thereof, e.g., obtained via predicate abstraction. The $\Gamma$-complete automata we define use a fixed set $D=\{a,b,c,a_1,\top\}$ of values ($a_1$ is needed to deal with the second item in Lemma~\ref{lemma:ob order has bounded length}). Therefore, for any $\Gamma$, $\mathcal{C}\cap A(\Gamma)\neq\emptyset$ iff $\mathcal{C}_D\cap A(\Gamma)\neq\emptyset$, where $\mathcal{C}_D$ is the subset of $\mathcal{C}$ that uses only values in $D$.

The set of executions $\mathcal{C}_D$ can be represented by a Vector Addition System with States (VASS). Since the set of values and priorities is bounded, each method invocation can be represented by a finite-state automaton (see~\cite{conf/esop/BouajjaniEEH13}). For a fixed set of priorities $P\subseteq \mathbb{P}$, the register automata $A(\Gamma)$ can be transformed to finite-state automata (the number of valuations of the registers is bounded). Thus, checking linearizability of an implementation in $\mathcal{C}$ is PSPACE when the number of threads is bounded, and EXPSPACE, otherwise.

%The set of executions $\mathcal{C}_D$ can be represented by a Vector Addition System with States (VASS). Since the set of values and priorities is bounded, each method invocation can be represented by a finite-state automaton (see~\cite{conf/esop/BouajjaniEEH13}). For a fixed set of priorities $P\subseteq \mathbb{P}$, the register automata $A(\Gamma)$ can be transformed to finite-state automata (the number of valuations of the registers is bounded). A VASS has a finite set of states, and manipulates a finite set of counters holding non-negative values. The idea is that the states of the VASS is a product of the valuation of global variables of $\mathcal{C}_D$ and state of $A(\Gamma)$. This set is finite. Each counter of the VASS then represents the number of threads which are at a particular control location within a method, with a certain valuation of the local variables. Here again, as the data values are bounded, there is a finite number of valuations. When a thread moves from a control location to another, we decrement, resp., increment, the counter corresponding the old, resp., the new control location. In this way, we model $\mathcal{C}_D$ as VASS. Thus, checking linearizability of an implementation in $\mathcal{C}$ is PSPACE when the number of threads is bounded, and EXPSPACE, otherwise.

%We model the set $\mathcal{C}_D$ as the executions of a Vector Addition System with States (VASS) similarly to Bouajjani et al. \cite{conf/esop/BouajjaniEEH13}. A VASS has a finite set of states, and manipulates a finite set of counters holding non-negative values. The idea is that the states of the VASS represent the global variables of $\mathcal{C}_D$. This set is finite as we have bounded the data values. Each counter of the VASS then represents the number of threads which are at a particular control location within a method, with a certain valuation of the local variables. Here again, as the data values are bounded, there is a finite number of valuations. When a thread moves from a control location to another, or updates its local variables, we decrement, resp., increment, the counter corresponding the old, resp., the new, control location or valuation.

%For a fixed set of priorities $P$, the register automata $A(\Gamma)$ can be transformed to finite-state automata since the number of possible valuations of the registers is bounded. In this way, we reduce checking linearizability of priority queue into the the EXPSPACE-complete problem of checking state-reachability in a VASS. When we consider only finite number of threads, the value of counters are bounded and then this problem is PSPACE \cite{DBLP:conf/ac/Esparza96}.

\noindent {\bf Reducing State-Reachability Problem of VASS into Linearizability Problem of Priority Queue:} We show how to reduce the state-reachability problem of VASS into checking linearizability of priority queue. Our construction is based on the reduction of state-reachability problem of counter machine and VASS into linearizability problem in \cite{conf/esop/BouajjaniEEH13}.

%There are already several linearizable lock-free implementations of priority queue \cite{DBLP:conf/ipps/SundellT03,DBLP:conf/ppopp/LiuS12}. It is not hard to modify them to fit finite variables with finite data domain and it is obvious that they are still linearizable then. Let $\textit{Imp-PQ}_{\textit{lf}}$ be one such implementations. It is safe to assume that $\textit{Imp-PQ}_{\textit{lf}}$ is given in the form of pseudo-code. We use $\textit{put}_{\textit{lf}}$ and $\textit{rm}_{\textit{lf}}$ to explicitly denote the pseudo-code of put and remove method of $\textit{Imp-PQ}_{\textit{lf}}$.

Let $\textit{Imp-PQ}_{\textit{lf}}$ be a correct priority queue implementation where (1) each method execute in a sequentially way (can be achieved by using lock and unlock) and (2) uses finite variables with finite data domain. It is quite easy to construct $\textit{Imp-PQ}_{\textit{lf}}$ and we omit its detailed definition here. By condition (1) we know that each method does not to wait for other method to complete its execution. It is safe to assume that $\textit{Imp-PQ}_{\textit{lf}}$ is given in the form of pseudo-code. We use $\textit{put}_{\textit{lf}}$ and $\textit{rm}_{\textit{lf}}$ to explicitly denote the pseudo-code of put and remove method of $\textit{Imp-PQ}_{\textit{lf}}$.

@@

%In \cite{conf/esop/BouajjaniEEH13}, to use a library to simulate counter machine $\mathcal{A}$, their library $\mathcal{L}_{\mathcal{A}}$ contains the following variables: (1) variable $q$ that uses value from $Q_{\mathcal{A}}$, (2) vector $\textit{req}[]$ and $\textit{ack}[]$, the index is chosen from $1$ to the number of unit vectors of length $s$, (3) $\textit{dec}[]$, where the index is chosen form $1$ to $s$.
